---
title: "ANOVA `desenvolvimento` ~ `area.de.conhecimento`"
author: Geiser C. Challco <geiser@usp.br>
comment: This file is automatically generate by Shiny-Statistic app (https://statistic.geiser.tech/)
         
         Shiny-Statistic is distributed in the hope that it will be useful,
         but WITHOUT ANY WARRANTY; without even the implied warranty of
         MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
         GNU General Public License for more details.
         
         You should have received a copy of the GNU General Public License.
         If not, see <https://www.gnu.org/licenses/>.
output:
  pdf_document:
    keep_tex: true
  github_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
wants <- c('ggpubr', 'emmeans', 'rstatix', 'fBasics', 'car', 'carData','dplyr','knitr')
has <- wants %in% rownames(installed.packages())
if (any(!has)) install.packages(wants[!has])

library(knitr)
opts_chunk$set(echo = TRUE)

library(dplyr)
library(carData)
library(car)
library(fBasics)
library(rstatix)
library(emmeans)

library(ggpubr)
library(moments)

library(stats)

normality_test <- function(x) {
  plimit <- 0.05
  n.test <- shapiro.test(x)
  cutpoints <- c(0, 1e-04, 0.001, 0.01, 0.05, 1)
  
  if (length(x) > 30) {
    plimit <- 0.01
    cutpoints <- c(0, 1e-05, 1e-04, 0.001, 0.01, 1)
  }
  if (length(x) > 50) n.test <- dagoTest(x)@test
  
  normality <- ifelse(n.test$p.value[1] < plimit, 'NO', 'YES')
  if (length(x) > 100) normality <- 'QQ'
  if (length(x) > 200) normality <- '-'
  
  return(cbind(add_significance(data.frame(
    n = length(x),
    statistic = n.test$statistic[1],
    method = strsplit(n.test$method, ' ')[[1]][1],
    p = n.test$p.value[1]
  ), p.col = "p", cutpoints = cutpoints), normality = normality))
}

normality_test_at <- function(dat, vars) {
  df <- select(group_data(dat), -starts_with(".rows"))
  do.call(rbind, lapply(vars, FUN = function(v) {
    do.call(rbind, lapply(seq(1, nrow(group_data(dat))), FUN = function(i) {
      n.test <- normality_test(dat[[v]][group_data(dat)[[".rows"]][[i]]]) 
      cbind(variable = v, df[i,] , n.test)
    }))
  }))
}

df2qqs <- function(data, group) {
  data <- as.data.frame(data)
  for (iv in group) {
    if (is.numeric(data[[iv]])) {
      quantiles <- quantile(data[[iv]])
      data[[iv]] <- sapply(data[[iv]], FUN = function(x) {
        if (x <= quantiles[[2]]) "low"
        else if (x >= quantiles[[4]]) "high"
        else "medium"
      })
      data[[iv]] <- factor(data[[iv]], levels=c("low", "medium", "high"))
    }
  }
  return(data)
}

ggPlotAoV <- function(data, x, y, color = c(), aov, pwc, linetype = color, by = c(), addParam = c() ) {
  pwc <- tryCatch(add_xy_position(pwc, x = x), error = function(e) NULL)
  if (is.null(pwc)) return(ggplot())
  if (length(color) > 0) {
    bxp <- ggboxplot(data, x = x, y = y, color = color, palette = "jco", add=addParam, facet.by = by)
    bxp <- bxp + stat_pvalue_manual(pwc, color = color, linetype = linetype, hide.ns = T
                                    , tip.length = 0, step.increase = 0.1, step.group.by = by)
  } else {
    bxp <- ggboxplot(data, x = x, y = y, color = x, palette = "jco", add=addParam, facet.by = by)
    bxp <- bxp + stat_pvalue_manual(pwc, linetype = linetype, hide.ns = T
                                    , tip.length = 0, step.increase = 0.1, step.group.by = by)
  }
  bxp <- bxp + labs(subtitle = get_test_label(aov, detailed = T), caption = get_pwc_label(pwc))
  return(bxp)
}

emm <- list()

dat <- read.csv("data.csv")[,c("ID", "area.de.conhecimento", "desenvolvimento" )]
dat <- df2qqs(dat, c("area.de.conhecimento"))
rownames(dat) <- dat[["ID"]]
```

## Initial Data and Preprocessing

R script: [factorialAnova.R](factorialAnova.R)
Inital data: [data.csv](data.csv)



### Summary statistics of the initial data

```{r}
get_summary_stats(group_by(dat, `area.de.conhecimento`), type ="common")
```

## Check Assumptions

### Identifying outliers

Outliers tend to increase type-I error probability, and they decrease the calculated F statistic in ANOVA resulting in a lower chance of reject the null hypothesis.

* Identified outliers using rstatix

```{r}
identify_outliers(group_by(dat, `area.de.conhecimento`), `desenvolvimento`)
```

* Identified outliers through Boxplots

```{r}
Boxplot(`desenvolvimento` ~ `area.de.conhecimento`, data = dat, id = list(n = Inf))
```

### Removing outliers from the data

```{r}
outliers <- c("Obs322")
rdat <- dat[!dat[["ID"]] %in% outliers,]   # table without outliers
```

```{r echo=FALSE}
kable(dat[dat[["ID"]] %in% outliers,], caption = 'Outliers table')
```

### Normality assumption

**Observation**:

As sample sizes increase, ANOVA remains a valid test even with the violation of normality <sup>[[1](#references), [2](#references)]</sup>.
According to the central limit theorem, the sampling distribution tends to be normal if the sample is large enough (`n > 30`). Therefore, we performed ANOVA with large samples as follows: 

- In cases with the sample size greater than 30 (`n > 30`), we adopted a significance level of `p < 0.01` instead a significance level of `p < 0.05`.

- For samples with `n > 50` observation, we adopted D'Agostino-Pearson test that offers better accuracy for larger samples <sup>[[3](#references)]</sup>.

- For samples' size between `n > 100` and `n <= 200`, we ignored both tests (Shapiro and D'Agostino-Persons), and our decision of normality were based only in the interpretation of QQ-plots and histograms because these tests tend to be too sensitive with values greater than 200 <sup>[[3](#references)]</sup>.

- For samples with `n > 200` observation, we ignore the normality assumption based on the central theorem limit, and taking only into account the homogeneity assumption.

#### Checking normality assumption in the residual model

```{r}
mdl <- lm(`desenvolvimento` ~ `area.de.conhecimento`, data = rdat)
normality_test(residuals(mdl))
```

The QQ plot used to evaluate normality assumption

```{r}
qqPlot(residuals(mdl))
```

#### Checking normality assumption for each group

```{r}
normality_test_at(group_by(rdat, `area.de.conhecimento`), "desenvolvimento")
```

* QQ plot in the **area.de.conhecimento**: "Ciências Agrárias" 
```{r}
qqPlot( ~ `desenvolvimento`, data = rdat[which(rdat["area.de.conhecimento"] == "Ciências Agrárias"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências Biológicas" 
```{r}
qqPlot( ~ `desenvolvimento`, data = rdat[which(rdat["area.de.conhecimento"] == "Ciências Biológicas"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências da Saúde" 
```{r}
qqPlot( ~ `desenvolvimento`, data = rdat[which(rdat["area.de.conhecimento"] == "Ciências da Saúde"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências Exatas e da Terra" 
```{r}
qqPlot( ~ `desenvolvimento`, data = rdat[which(rdat["area.de.conhecimento"] == "Ciências Exatas e da Terra"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências Humanas" 
```{r}
qqPlot( ~ `desenvolvimento`, data = rdat[which(rdat["area.de.conhecimento"] == "Ciências Humanas"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências Sociais Aplicadas" 
```{r}
qqPlot( ~ `desenvolvimento`, data = rdat[which(rdat["area.de.conhecimento"] == "Ciências Sociais Aplicadas"),])
```

* QQ plot in the **area.de.conhecimento**: "Engenharias" 
```{r}
qqPlot( ~ `desenvolvimento`, data = rdat[which(rdat["area.de.conhecimento"] == "Engenharias"),])
```

* QQ plot in the **area.de.conhecimento**: "Linguística/Letras e Artes" 
```{r}
qqPlot( ~ `desenvolvimento`, data = rdat[which(rdat["area.de.conhecimento"] == "Linguística/Letras e Artes"),])
```



#### Removing data that affect normality

```{r}
non.normal <- c("")
sdat <- rdat[!rdat[["ID"]] %in% non.normal,]   # table without non-normal and outliers
```

```{r echo=FALSE}
kable(rdat[rdat[["ID"]] %in% non.normal,], caption = 'Non-normal data table')
```

#### Performing normality test without data that affect normality 

```r
mdl <- lm(`desenvolvimento` ~ `area.de.conhecimento`, data = sdat)
normality_test(residuals(mdl))
```

```{r echo=FALSE}
mdl <- lm(`desenvolvimento` ~ `area.de.conhecimento`, data = sdat)
kdf <- normality_test(residuals(mdl))
kdf$p <- round(kdf$p, 4)
kdf$p[which(kdf$p < 0.0001)] <- '< 0.0001'
kable(kdf, digits = 4)
```

```r
normality_test_at(group_by(sdat, `area.de.conhecimento`), "desenvolvimento")
```

```{r echo=FALSE}
kdf <- add_significance(normality_test_at(group_by(sdat, `area.de.conhecimento`), "desenvolvimento"))
kdf$p <- round(kdf$p, 4)
kdf$p[which(kdf$p < 0.0001)] <- '< 0.0001'
kable(kdf, digits = 4)
```

QQ plot in the residual model without data that affect normality

```{r}
qqPlot(residuals(mdl))
```

* QQ plot in the **area.de.conhecimento**: "Ciências Agrárias" 
```{r}
qqPlot( ~ `desenvolvimento`, data = sdat[which(sdat["area.de.conhecimento"] == "Ciências Agrárias"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências Biológicas" 
```{r}
qqPlot( ~ `desenvolvimento`, data = sdat[which(sdat["area.de.conhecimento"] == "Ciências Biológicas"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências da Saúde" 
```{r}
qqPlot( ~ `desenvolvimento`, data = sdat[which(sdat["area.de.conhecimento"] == "Ciências da Saúde"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências Exatas e da Terra" 
```{r}
qqPlot( ~ `desenvolvimento`, data = sdat[which(sdat["area.de.conhecimento"] == "Ciências Exatas e da Terra"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências Humanas" 
```{r}
qqPlot( ~ `desenvolvimento`, data = sdat[which(sdat["area.de.conhecimento"] == "Ciências Humanas"),])
```

* QQ plot in the **area.de.conhecimento**: "Ciências Sociais Aplicadas" 
```{r}
qqPlot( ~ `desenvolvimento`, data = sdat[which(sdat["area.de.conhecimento"] == "Ciências Sociais Aplicadas"),])
```

* QQ plot in the **area.de.conhecimento**: "Engenharias" 
```{r}
qqPlot( ~ `desenvolvimento`, data = sdat[which(sdat["area.de.conhecimento"] == "Engenharias"),])
```

* QQ plot in the **area.de.conhecimento**: "Linguística/Letras e Artes" 
```{r}
qqPlot( ~ `desenvolvimento`, data = sdat[which(sdat["area.de.conhecimento"] == "Linguística/Letras e Artes"),])
```



### Homogeneity of variance assumption

```r
levene_test(sdat, `desenvolvimento` ~ `area.de.conhecimento`)
```

```{r echo=FALSE}
kdf <- add_significance(levene_test(sdat, `desenvolvimento` ~ `area.de.conhecimento`))
kdf$p <- round(kdf$p, 4)
kdf$p[which(kdf$p < 0.0001)] <- '< 0.0001'
kable(kdf, digits = 4)
```

From the output above, non-significant difference indicates homogeneity of variance in the different groups (Signif. codes:  0 **** 0.0001 *** 0.001 ** 0.01 * 0.05 ns 1).

## Computation ANOVA

```r
res.aov <- anova_test(sdat, `desenvolvimento` ~ `area.de.conhecimento`, type = 2, effect.size = 'ges', detailed = T)
get_anova_table(res.aov)
```

```{r echo=FALSE}
res.aov <- anova_test(sdat, `desenvolvimento` ~ `area.de.conhecimento`, type = 2, effect.size = 'ges', detailed = T)
kdf <- get_anova_table(res.aov)
kdf$p <- round(kdf$p, 4)
kdf$p[which(kdf$p < 0.0001)] <- '< 0.0001'
kable(kdf, digits = 4)
```

## Post-hoct Tests (Pairwise Comparisons)

* Estimated marginal means for **area.de.conhecimento**
```r
(emm[["area.de.conhecimento"]] <- emmeans_test(sdat, `desenvolvimento` ~ `area.de.conhecimento`, p.adjust.method = "bonferroni", detailed = T))
```
```{r echo=FALSE}
emm[["area.de.conhecimento"]] <- emmeans_test(sdat, `desenvolvimento` ~ `area.de.conhecimento`, p.adjust.method = "bonferroni", detailed = T)
kdf <- add_significance(emm[["area.de.conhecimento"]])
kdf$p.adj <- round(kdf$p.adj, 4)
kdf$p.adj[which(kdf$p.adj < 0.0001)] <- '< 0.0001'
kable(kdf, digits = 4)
```



## Descriptive Statistic and ANOVA Plots 

```r
get_summary_stats(group_by(sdat, `area.de.conhecimento`), type ="common")
```

```{r echo=FALSE}
kdf <- get_summary_stats(group_by(sdat, `area.de.conhecimento`), type ="common")
cnames <- c("n","mean","median","min","max","sd","se","ci","iqr")
kdf <- kdf[,c(colnames(kdf)[!colnames(kdf) %in% cnames], cnames)]
kable(kdf, digits = 4)
```

```{r, fig.width=20, fig.height=10}
ggPlotAoV(sdat, "area.de.conhecimento", "desenvolvimento", aov=res.aov, pwc=emm[["area.de.conhecimento"]], addParam=c("jitter"))
```



## References

<sup>[1]</sup>: Blanca, M. J., Alarcón, R., Arnau, J., Bono, R., & Bendayan, R. (2017). Non-normal data: Is ANOVA still a valid option?. Psicothema, 29(4), 552-557.

<sup>[2]</sup>: Ghasemi, A., & Zahediasl, S. (2012). Normality tests for statistical analysis: a guide for non-statisticians. International journal of endocrinology and metabolism, 10(2), 486.

<sup>[3]</sup>: Miot, H. A. (2017). Assessing normality of data in clinical and experimental trials. J Vasc Bras, 16(2), 88-91.